Paper_Notes

### 2017-12-03 Paper_Notes.Rmd
Paper notes assignment for 2017-12-4

>>>>

### Hung et al. 2011

- Hung, Yang, Hu, Weng, and DeLisi. “Gene set enrichment analysis: performance evaluation and usage guidelines.” Briefings in Bioinformatics 13, no. 3, (2011): 281 – 91.

The article describes analytical methods for identifying related sets of genes associated with phenotypic changes (i.e. stage of differentiation, disease state, responsiveness to outside stimuli, etc.). Early methods for associating gene sets with phenotype changes first identify individual, potentially relevant genes by making a binary decision based on a quantity that measures the extent of differential expression between the phenotypes and then use a Fisher’s exact test to determine whether a significant number of these genes belong to a predetermined gene set. An alternative approach, named Gene Set Enrichments Analysis (GSEA), ranks all genes according to differential expression, and then determines if a predetermined gene set is significantly overrepresented. Performing GSEA is complex because of the numerous variants that are dependent on the method for estimating significance, background distribution, choice of the shuffling method, the method of multiple hypothesis correction, and the choice of weights to account for auxiliary information. The lack of a gold standard has greatly hampered the effort of assessing gene set enrichment methods using experimental data sets. Simulated data cannot substitute experimental data, as biological systems are so complex. The authors use GSEA to analyze 132 experimental data sets to assess six gene set-level statistics and one global test. The results presented in the paper suggest that the Wilcoxon rank sum test and the WKS test as implemented in GSEA provide the most effective gene set-level statistics for obtaining high mutual coverage and incorporating topology of gene sets in the analysis increases the sensitivity for all procedures. Furthermore, simulated background distributions are more accurate than analytic backgrounds and the mean and median tests achieve the highest sensitivity but poor mutual coverage.

### Bard et al. 2004

- Bard and Rhee. “Ontologies in Biology: Design, Applications and Future Challenges.” Nature Reviews Genetics, 5, (2004): 213 – 22.

Ontologies are a formal way of representing knowledge in which concepts are described both by their meaning and by their relationship to each other. Biological based ontologies (so called bio-ontologies) can be used for linking to and querying molecular databases. The article reviews the principal bio-ontologies and the current issues in their design and development. Ontologies are different from annotations in that they formalize the meaning of terms through a set of assertions and rules that are collectively known as a description logic. An advantage of ontologies is that the description logic can be used for querying an information set and for facilitating analyses across information sets that are not traditionally accessible to searching and comparing. Ontologies have been used in biology for some time. The Gene Ontology (GO) is by far the most widely used bio-ontology. It aims to formalize our knowledge about biological processes, molecular functions and cell components, in three orthogonal hierarchies. At the time of publication of the article, Cell Ontology was a new and still unfinished ontology, which was designed to provide all model species with a common language and ID set for cell phenotypes. These bio-ontologies have a wide variety of uses, the most important of which is the representation of knowledge in a computer-comprehensible way, interoperability across databases, and the annotation and analysis of large-scale data with ontology IDs. However, integrating this knowledge poses two problems. First, not everyone in a field agrees on either the facts or the relationships. Second, knowledge changes with time, even in apparently ossified subjects such as anatomy. If ontologies are properly curated over the longer term they will come to be seen as modern-day textbooks providing online and up-to-date biological expertise for their area.
